---
title: "Week 1: Assignment 1"
author: "Dominique Mashini"
date: "07-09-2020"
output: html_document
---

# Creating a database based on census data (Chile)

## Introduction: survey and sample

I decided to work with the Chilean survey National Study of Public Opinion N°84 (December, 2019), which includes data related to life satisfaction. 

The unit of analysis is at the person-level, and the sample includes 1,496 people surveyed at home in 117 counties across the country. The study includes a population aged over 18 years (urban and rural) residing throughout the country. The instrument was in person (face-to-face) and the interviews were based on a structured questionnaire. 


## Reading packages
In order to read the data I added the tidyverse and haven libraries.

```{r}
library(tidyverse)
library(haven)
```

## Unziping and reading data in SPSS format

I downloaded the data from CEP Chile (2019): https://www.cepchile.cl/cep/encuestas-cep/encuestas-2009-2018/estudio-nacional-de-opinion-publica-n-84-diciembre-2019 

```{r}

unzip("encuesta_cep84_dic2019.zip")

```

I named the full downloaded dataset data_cep.

```{r}

data_cep <- read_sav("Encuesta CEP 84 Dic 2019 v1.sav")

```

# Variables of the dataset

## Identification of variables

To identify the variables of interest, I reviewed the survey report and the survey research manual in the following link: https://www.cepchile.cl/cep/site/docs/20200116/20200116081636/encuestacep_diciembre2019.pdf .

The data contains 1496 observations of 258 variables.

## Exploring variables

After a general overview of the survey results, I found interesting to explore the outcome of car ownership based on certain variables collected by the survey. The variables describe 2 categorical variables (sex and car ownership) and 3 continuous variables (age, average weekly working hours, and years of formal education).

## Filtering dataset with only variables of interest

I filtered my dataset into 5 variables with 1496 observations defined in a dataset named data_cep_2.
In this process, I selected the following variables of interest:

1) DS_P2_EXACTA = AGE (continuous)
2) DS_P1 = SEX (categorical)
3) DS_P37_1 = CAR OWNERSHIP (categorical)
4) DS_P6_OTRO = AVERAGE WEEKLY WORKING HOURS (continuous)
5) DS_P3 = YEARS OF FORMAL EDUCATION (continuous)

```{r}
data_cep_2 <- select(data_cep, DS_P2_EXACTA, DS_P1, DS_P37_1, DS_P6_OTRO, DS_P3)
```
 
## Reviewing value labels of each variable.

For a better understanding of the dataset and the variables, I browsed through the results of the 1,496 observations and identified the labels related to the dataframe in each categorical variable using the function "data_cep_2$VARIABLEID". 

I found the following labels:
SEX (1= MALE, 2= FEMALE)

CAR OWNERSHIP (1=YES, 2=NO, 8=DOES NOT KNOW, 9=DOES NOT ANSWER)


Finally, I will save both datasets as CSV files, so I can use them later.
The first dataset corresponds to the full survey with the 258 variables.

```{r}
write_csv(data_cep, "data_cep.csv")
```

The second dataset corresponds to the filtered survey with my 5 variables of interest.
```{r}
write.csv(data_cep_2, "data_cep_2.csv")
```


