---
title: "Assignment3"
author: "Dominique Mashini"
date: "19-09-2020"
output: html_document
html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
    
---

# Assignment 3: Relationships between variables

## Loading packages for the assignment

First, I will load the following libraries for this assignment:

```{r load libraries, message=FALSE}
library(ggplot2)
library(tidycensus)
library(tidyverse)
```

```{r load themes, message= FALSE}
library(ggthemes)
```

## Reading and filtering the data

I obtained the data from the 1-year ACS Public Use survey at the person-level in the state of Texas (year 2018).
I created a dataframe with the list of all the variables available from the census package. I filtered the variables with the following function:

```{r}
person_variables_2018 <- 
  pums_variables %>% 
  distinct(year, survey, var_code, var_label, data_type, level) %>% 
  filter(level == "person", year == 2018, survey == "acs1")
```

## Defining, combining, and renaming variables

After browsing through the database person_variables_2018, I defined and renamed my 5 variables of interest:
	
1. Travel time to work (var code: JWMNP) = Traveltime (continuous)
2. Age (var code: AGEP) = Age (continuous)
3. Educational attainment (var code: SCHL) = Edattain (categorical)
4. Means of Transportation (var code: JWTR) = Meanstransp (categorical)
5. Total person's income (var code: PINCP) = Income (continuous)

I filtered the data, eliminating the observations with 0 and negative values in the income and travel time variables. I also selected my variables of interest for my categorical variables.
I merged some sub-categories of means of transportation, because it is interesting to analyze non-motorized mobility (bicycle or walked), public transportation (bus, streetcar, or subway), and private transportation (car, truck, or van). 

```{r variables, message = FALSE, results='hide'}

vars <- c(
  Traveltime = "JWMNP",
  Age = "AGEP",
  Edattain = "SCHL",
  Meanstransp = "JWTR",
  Income = "PINCP"
)

data <- get_pums(
  variables = vars,
  state = "TX",
  year = 2018,
  survey = "acs1",
  recode = TRUE
) %>%
    filter(Traveltime > 1) %>%
  filter(Income > 0) %>%
  filter(Meanstransp == "01"| Meanstransp == "02"| Meanstransp == "03"|Meanstransp == "04"|Meanstransp == "09"|Meanstransp == "10") %>%
  filter(Edattain == "01"| Edattain == "16" | Edattain == "19" | Edattain == "21" | Edattain == "22" | Edattain == "24") %>%
mutate(Meanstransp = case_when(
    Meanstransp == "01" ~ "Car, truck, or van",
    Meanstransp == "03" | Meanstransp == "04"  | Meanstransp == "02" ~ "Bus, streetcar or subway",
    Meanstransp == "09" | Meanstransp == "10" ~ "Bicycle or walked")) %>%

  mutate(Edattain = case_when(
    Edattain == "01" ~ "1 No schooling completed",
    Edattain == "16" ~ "2 Highschool diploma",
    Edattain == "19" ~ "3 College credit",
    Edattain == "21" ~ "4 Bachelor",
    Edattain == "22" ~ "5 Masters",
    Edattain == "24" ~ "6 Doctorate"))
    

```


# Relationships between variables

I want to explore the relationship between age, educational attainment, travel time, means of transportation and income on the person level. 


## Correlation between Continuous Variables

First, I will test the relationship between my continuous variables (age, travel time and income) by running a correlation test.


```{r}
correlation1 <- cor.test(data$Age, data$Traveltime)

correlation1
```




```{r}
correlation2 <- cor.test(data$Age, data$Income)

correlation2
```

```{r}
correlation3 <- cor.test(data$Traveltime, data$Income)

correlation3
```
Correlation analysis based on Pearson's r and p-values:

In my three correlation analysis I see a weak, but statistically significant, positive relationship at a 95% confidence level. The stronger correlation is seen between age and income (cor =0.24).
To visualize these relationships, I will use scatter plot graphs:


```{r scatterplot, message= FALSE}
ggplot(data, aes(x = Age, y = Traveltime)) + 
    geom_point(size = 0.5, alpha = 0.25, position = "jitter") +
    geom_smooth() +
    theme_economist_white() +
    scale_x_continuous(name = "Age (years)") +
    scale_y_continuous(name = "Travel time (minutes)",
                      breaks = seq(0, 150, by = 10), 
                      labels = paste("",
                                  prettyNum(seq(0, 150, by = 10),
                                            big.mark = ","))) +
      annotate(geom = "text", x = 16, y = 180, 
           label = paste("Correlation (95-percent confidence):",
                         prettyNum(correlation1$conf.int[1], digits = 2), 
                         "to",
                         prettyNum(correlation1$conf.int[2], digits = 2)),
           hjust = 0)
```


Looking at the scatter plot graph, it is more evident that the correlation is weaker and slightly negative over 65 years.


```{r scatterplot2, message= FALSE}
ggplot(data, aes(x = Age, y = Income)) + 
    geom_point(size = 0.4, alpha = 0.2, position = "jitter") +
    geom_smooth() +
    theme_economist_white() +
    scale_x_continuous(name = "Age (years)") +
    scale_y_continuous(name = "Income (USD)",
                      breaks = seq(0, 1000000, by = 100000), 
                      labels = paste("",
                                  prettyNum(seq(0, 1000000, by = 100000),
                                            big.mark = ","))) +
          annotate(geom = "text", x = 20, y = 1100000, 
           label = paste("Correlation (95-percent confidence):",
                         prettyNum(correlation2$conf.int[1], digits = 2), 
                         "to",
                         prettyNum(correlation2$conf.int[2], digits = 2)),
           hjust = 0)
```

I can see a stronger, positive correlation between age an income. 

```{r scatterplot3, message= FALSE}
ggplot(data, aes(x = Income, y = Traveltime)) + 
    geom_point(size = 0.5, alpha = 0.2, position = "jitter") +
    geom_smooth() +
    theme_economist_white() +
    scale_x_continuous(name = "Income (USD)") +
    scale_y_continuous(name = "Travel Time (minutes)",
                      breaks = seq(0, 150, by = 10), 
                      labels = paste("",
                                  prettyNum(seq(0, 150, by = 10),
                                            big.mark = ","))) +
       annotate(geom = "text", x = 0, y = 170, 
           label = paste("Correlation (95-percent confidence):",
                         prettyNum(correlation3$conf.int[1], digits = 2), 
                         "to",
                         prettyNum(correlation3$conf.int[2], digits = 2)),
           hjust = 0)
```


There is an interesting visualization of the correlation between income and travel time. While positive, there are fewer high-income people who travel more than 150 minutes. 


## Relationship between Categorical Variables

I will use the Chi-square test of association for determining whether there is a statistically significant association between my two categorical variables.


```{r}
chi_sq <- chisq.test(data$Edattain, data$Meanstransp)
chi_sq
```



```{r}
knitr::kable(chi_sq$observed)
```



```{r}
knitr::kable(chi_sq$expected)
```



```{r}
C_diff <- chi_sq$observed - chi_sq$expected

C_diff_data <- as.data.frame(C_diff)
C_diff_data
```


I can assume there is a statistically significant relationship at a 95% confidence interval between educational attainment and means of transportation because the p-value is below 0.05. 

In the tables I see some values that are particularly low, such as no schooling completed with non-motorized mobilization and doctorate with bus, streetcar or subway.



## Relationships between Educational Attainment and Continuous Variables

For the relationship between my categorical and continuous variables, I will use the anova test:

```{r}
anova <- aov(data$Age ~ data$Edattain)

summary(anova)
```

```{r}
anova <- aov(data$Traveltime ~ data$Edattain)

summary(anova)
```
```{r}
anova <- aov(data$Income ~ data$Edattain)

summary(anova)
```
Difference Check:

```{r}
differences <- TukeyHSD(anova)
differences
```

There exists a significant association between educational attainment and age, income and travel time. On a more in-depth review of income and educational attainment, a positive difference and significant association appear, except between highschool diploma and no schooling completed (p adj = 0.085).

I decided to display the relationship between educational attainment and income using a box-plot. As my categorical variable is ordinal, I ordered the educational attainment categories for a better visualization of the relationship:


```{r}
ggplot(data, aes(x = Edattain, y = Income)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(name = "Educational Attainment") +
  scale_y_continuous(name = "Income (USD)",
                     breaks = seq(0, 1000000, by = 100000),
                     labels = paste(prettyNum(seq(0, 1000000, by = 100000), 
                                              big.mark = ",")))
```

## Relationships between Means of Transportation and Continuous Variables

For the relationship between my categorical and continuous variables, I use the anova test:


```{r}
anova <- aov(data$Age ~ data$Meanstransp)

summary(anova)
```

```{r}
anova <- aov(data$Traveltime ~ data$Meanstransp)

summary(anova)
```

```{r}
anova <- aov(data$Income ~ data$Meanstransp)

summary(anova)
```
```{r}
differences <- TukeyHSD(anova)
differences
```


There exists a significant relationship between means of transportation and all the continuous variables at a 95% confidence level, with all p-values under the 0.05 threshold.
In the difference check between income and means of transportation, there is a lower incidence of income between the variables 'bus, streetcar or subway' and 'bicycle or walked'.

I decided to display the relationship between travel time and means of transportation using a box-plot: 


```{r}
ggplot(data, aes(x = Meanstransp, y = Traveltime)) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(name = "Means of Transportation") +
  scale_y_continuous(name = "Travel Time (minutes)",
                     breaks = seq(0, 200, by = 20),
                     labels = paste(prettyNum(seq(0, 200, by = 20), 
                                              big.mark = ",")))
```



As expected, the travel time is longer and more varied in the 'bus, streetcar, or subway' variable, followed by 'car, truck, or van'. The time traveled by bicycle or walked is less.


## Relationship between Means of Transportation (Motorized and Non-motorized) and Age

To further my analysis, I combined the categories of means of transportation into motorized and non-motorized, in order to test them with age through a two-sample t-test. 


```{r}
data1 <- data %>% 
    mutate(
        motor = case_when(
            Meanstransp == "Car, truck, or van" |  Meanstransp == "Bus, streetcar or subway" ~ "motor",
            Meanstransp == "Bicycle or walked"  ~ "no motor"
        ))

differences1 <- t.test(data1$Age ~ data1$motor == "motor")
differences1
```

There is a significant relationship (negative, seen in the 95 percent confidence interval) found between age and means of transportation (motorized and non-motorized); the youngest use more non-motorized transport.
